---
title: "Mobilome"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
library(readr)
library(dplyr)
library(vegan)
library(ape)
library(phytools)
library(phylosignal)
library(nlme)
library(MASS)
```


```{r, warning=FALSE, message=FALSE}
bins_extendedinfo <- read_csv("~/Documents/UMich/papers/GrowthDefenseTradeOff/R output/bins_extendedinfo.csv")
allmags_cogs <- read_csv("~/Documents/UMich/papers/GrowthDefenseTradeOff/R output/cog_allmags.csv")
sigmags_cogs <- read_csv("~/Documents/UMich/papers/GrowthDefenseTradeOff/R output/expanded_df_cog.csv")
shared_cog_proportions <- read.csv("~/Documents/UMich/papers/GrowthDefenseTradeOff/R output/cog_proportions_shared.csv")
#shared_pfam_proportions <- read.csv("~/Documents/UMich/papers/GrowthDefenseTradeOff/R output/filtered_proportions_for_tests.csv")
#enriched_pfam <- read.csv("~/Documents/UMich/papers/GrowthDefenseTradeOff/R output/enriched_pfams.csv")
enriched_KEGG <- read.csv("~/Documents/UMich/papers/GrowthDefenseTradeOff/R output/kegg_merged.csv")
enriched_COG <- read.csv("~/Documents/UMich/papers/GrowthDefenseTradeOff/R output/final_results_cog.csv") 
```


```{r, warning=FALSE, message=FALSE}
# resistance trait value (quagga_mean/control_mean)
bins_extendedinfo <- bins_extendedinfo %>%
  mutate(resistance_trait = quagga_mean / control_mean)

# mean resistance trait across dates for each MAG
resistance_mean <- bins_extendedinfo %>%
  group_by(bin) %>%
  summarise(mean_resistance = mean(resistance_trait, na.rm = TRUE)) %>%
  ungroup()
```


```{r, warning=FALSE, message=FALSE}
# merge taxonomy
mags_taxinfo <- bins_extendedinfo %>% 
  dplyr::select(bin, t_phylum, t_class, t_order, t_family, t_genus, t_species) %>% 
  rename(bin_name = bin)

phyla_colors <- c(
  Proteobacteria = "#E1A91F",
  Bacteroidota = "#A6BF96",
  Verrucomicrobiota = "#F4B5C8",
  Planctomycetota = "#B39BC8",
  Unknown = "#D3D3D3", 
  Actinobacteriota = "#F4A259",
  Cyanobacteria = "#78C2BE", 
  Myxococcota = "#C7B198",
  Gemmatimonadota = "#8D8DAA",
  Chloroflexota = "#BCCF02"
)
```


```{r, warning=FALSE, message=FALSE}
mobilome_cogs <- sigmags_cogs %>% 
  dplyr::filter(function. == "Mobilome: prophages, transposons")
```


```{r, warning=FALSE, message=FALSE}
# Richness
richness_df <- mobilome_cogs %>%
  dplyr::group_by(bin_name) %>%
  summarise(richness_mobilome = n_distinct(gene_callers_id)) %>%
  left_join(mobilome_cogs %>% dplyr::select(bin_name, group) %>% distinct(), by = "bin_name")


# Shannon Diversity
shannon_df <- mobilome_cogs %>%
  group_by(bin_name) %>%
  summarise(shannon_mobilome = vegan::diversity(table(gene_callers_id))) %>% 
  left_join(mobilome_cogs %>% dplyr::select(bin_name, group) %>% distinct(), by = "bin_name")

# Richness Wilcoxon test 
wilcox_richness <- wilcox.test(richness_mobilome ~ group, data = richness_df)

# Shannon Diversity Wilcoxon test
wilcox_shannon <- wilcox.test(shannon_mobilome ~ group, data = shannon_df)

# results
print("Richness Comparison:")
print(wilcox_richness)

print("Shannon Diversity Comparison:")
print(wilcox_shannon)
```
Mobilome diversity is significantly different between resistant and susceptible populations:
- Richness (p = 0.0074): Resistant and susceptible MAGs have significantly different numbers of unique mobilome-related genes. Higher richness in resistant MAGs suggests more diverse mobilome activity/function in resistant populations.
- Shannon Diversity (p = 0.0080): Higher in resistant MAGs here indicates greater mobilome evenness, further supporting the idea that mobilome-associated genes are may play an adaptive role in conferring resistance to mussel predation.

```{r, warning=FALSE, message=FALSE}
# phylum_info
mags_phy <- mags_taxinfo %>% 
  filter(bin_name %in% richness_df$bin_name) %>% 
  dplyr::select(bin_name, t_phylum) %>% 
  distinct() %>% 
    mutate(t_phylum = replace_na(t_phylum, "Unknown"))

# merge taxonomy info
richness_df <- left_join(richness_df, mags_phy, by = "bin_name")
shannon_df <- left_join(shannon_df, mags_phy, by = "bin_name")

# replace labels
richness_df <- richness_df %>% dplyr::mutate(group = case_when(group == "resistant" ~ "Resistant", 
                                         group == "susceptible" ~ "Susceptible"))
shannon_df <- shannon_df %>% dplyr::mutate(group = case_when(group == "resistant" ~ "Resistant", 
                                         group == "susceptible" ~ "Susceptible"))

custom_colors <- c("Resistant" = "#B22222", "Susceptible" = "#808080")
```

Make plots comparing mobilome Richness and Shannon
```{r, warning=FALSE, message=FALSE}
Figure_3C_richness <- ggplot(richness_df, aes(x = group, y = richness_mobilome, fill = group)) +
  #geom_violin(alpha = 0.3, width = 0.8, color = NA) +  
  geom_boxplot(alpha = 0.9, width = 0.5, outlier.shape = NA) +  
  #geom_jitter(aes(color = group), width = 0.1, alpha = 0.5, size = 3) +  
  scale_y_log10() +  
  scale_fill_manual(values = custom_colors, name = "Population category") +  
  scale_color_manual(values = custom_colors, name = "Population category") +
  labs(title = "Mobilome Richness", y = "Richness (log10)", x = "Population category") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, vjust = 8, face = "bold", size = 22),
    plot.margin = ggplot2::margin(50, 50, 10, 10),  
    axis.title = element_text(size = 18, face = "bold"),
    axis.text.y = element_text(size = 14),  
    axis.text.x = element_text(size = 14),  
    axis.line = element_line(size = 0.8),
    legend.position = "none",
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14, face = "bold"),
    strip.text = element_text(size = 16, face = "bold"), 
    panel.spacing = unit(1.5, "lines"),  
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    panel.background = element_rect(fill = "white")
  )

Figure_3C_shannon <- ggplot(shannon_df, aes(x = group, y = shannon_mobilome, fill = group)) +
  #geom_violin(alpha = 0.3, width = 0.8, color = NA) +  
  geom_boxplot(alpha = 0.9, width = 0.5, outlier.shape = NA) +  
  #geom_jitter(aes(color = group), width = 0.2, alpha = 0.8, size = 3.5) +  
  scale_fill_manual(values = custom_colors, name = "Population category") +  
  scale_color_manual(values = custom_colors, name = "Population category") +
  labs(title = "Mobilome Shannon Diversity", y = "Shannon index", x = "Population category") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, vjust = 8, face = "bold", size = 22),
    plot.margin = ggplot2::margin(50, 50, 10, 10),  
    axis.title = element_text(size = 18, face = "bold"),
    axis.text.y = element_text(size = 14),  
    axis.text.x = element_text(size = 14),  
    axis.line = element_line(size = 0.8),
    legend.position = "none",
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14, face = "bold"),
    strip.text = element_text(size = 16, face = "bold"), 
    panel.spacing = unit(1.5, "lines"),  
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    panel.background = element_rect(fill = "white")
  )

print(Figure_3C_shannon)
print(Figure_3C_richness)

ggsave(filename = "../R output/Figures/Figure_3C_richness.jpg", 
       plot = Figure_3C_richness, 
       width = 7, height = 7, dpi = 300, units = "in")
ggsave(filename = "../R output/Figures/Figure_3C_shannon.jpg", 
       plot = Figure_3C_shannon, 
       width = 7, height = 7, dpi = 300, units = "in")
```

Correlation between KEGG and mobile diversity.
```{r, warning=FALSE, message=FALSE}
kegg_mobilome <- enriched_KEGG %>%
  #dplyr::select(-group) %>% 
  left_join(shannon_df, by = "bin_name") %>% 
  dplyr::select(-group.y) %>% 
  rename(group = "group.x")

# filter modules with zero completeness
kegg_filtered <- kegg_mobilome %>%
  filter(stepwise_module_completeness > 0)

# merge mobilome diversity and KEGG data
kegg_summary_shannon <- kegg_filtered %>%
  group_by(bin_name) %>%
  dplyr::summarise(kegg_count = n_distinct(module),  # unique KEGG module count
            mean_mobilome_shannon = mean(shannon_mobilome))  # mean Shannon diversity of mobilome

# check correlation
cor_test_shannon <- cor.test(kegg_summary_shannon$kegg_count, kegg_summary_shannon$mean_mobilome_shannon, method = "spearman")
print(cor_test_shannon)
```
This indicates that populations with higher mobilome Shannon diversity index tend to have a greater number of unique KEGG modules (Spearman's Ï = 0.36, p < 0.001).


```{r, warning=FALSE, message=FALSE}
groupinfo <- shannon_df %>% 
  dplyr::select(bin_name, group)
kegg_summary_shannon <- left_join(kegg_summary_shannon, groupinfo, by = "bin_name")

# Define custom legends
custom_colors <- c("Resistant" = "#B22222", "Susceptible" = "#808080")
custom_shapes <- c("Resistant" = 17, "Susceptible" = 16)

# Scatter plot of KEGG enrichment vs. Shannon diversity
Figure_S3 <- ggplot(kegg_summary_shannon %>% drop_na(mean_mobilome_shannon, kegg_count, group), 
                               aes(x = mean_mobilome_shannon, y = kegg_count)) +
  geom_point(aes(color = group), alpha = 0.7, size = 5) +
  geom_smooth(method = "lm", se = TRUE, color = "black", fill = "gray70", linewidth = 1) +
  scale_color_manual(values = custom_colors) +  # Colors (no legend)
  scale_shape_manual(values = custom_shapes, name = "Population Category") +
  labs(title = "KEGG Module Count vs. Mobile Elements Diversity (Shannon)",
       x = "Shannon Diversity of Mobile Elements", y = "Number of Enriched KEGG Modules") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, vjust = 1, face = "bold", size = 22),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text.y = element_text(size = 10),  
    axis.text.x = element_text(size = 14),
    axis.line = element_line(size = 0.8),
    legend.position = "right",
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14, face = "bold"),
    strip.text = element_blank(),
    panel.spacing = unit(3, "lines"),  
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  # Remove color legend and keep shape legend
  guides(color = "none")

print(Figure_S3)
ggsave(filename = "../R output/Figures/Figure_S3.jpg", 
       plot = Figure_S3, 
       width = 15, height = 10, dpi = 300, units = "in")
```


Run GLM
```{r, warning=FALSE, message=FALSE}
glm_kegg_shannon <- glm(kegg_count ~ mean_mobilome_shannon, 
                         family = poisson(link = "log"), 
                         data = kegg_summary_shannon)
summary(glm_kegg_shannon)
```



